A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-05-11 podc_paper/algorithm.tex
minor fixes

\subsection{Overview}
Our \emph{\ordering\ tree} data structure is used to agree on a total ordering of the operations performed on the queue.
It is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
for each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{orderingtree} for an example.
In this section, we use an infinite array of blocks in each node.
Section \ref{reducing} describes how to replace the infinite array by a representation with bounded space.

@@

\Eric{Reworded following parag after Hossein's comments of Jan 4}
To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ first observes
the blocks in both of $\var{v}$'s children that are not already in $\var{v}$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this a 3-step sequence a
\op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on $\var{v}$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $\var{v}$'s children to $\var{v}$,
we can prove that if two \op{Refresh}es by $P$ on $\var{v}$ fail,
then $P$'s operation has been propagated to $\var{v}$ by some other process, so $P$ can continue 
onwards towards the~root.

@@

\begin{figure*}[t]
\input{tournamentTree.pdf_t}
\caption{An example \ordering\ tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still propagating towards the root.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
\medskip
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{orderingtree}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as before.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure*}

@@

Performing a double refresh at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation completes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in the root's blocks array.
Thus, we linearize operations according to the block they belong to in the root's array.
We can choose how to order operations in the same block, since they must be concurrent.

@@

Finally, when linearizing the operations propagated to the root, we must
choose how to order operations within a block.  Any ordering within a block would work, since all operations
in a block are concurrent with one another.  We choose to put
each block's enqueues before its dequeues.
Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
linearization ordering is 
\begin{equation}
L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
\label{linearization}
\end{equation}

@@

The only remaining field is $B.\fld{super}$.  When the block 
$B$ is created for a \node\ $\var{v}$, we do not yet know where its
superblock will eventually be installed in $B$'s parent.
So, we leave $B.\fld{super}$ blank.  
Soon after $B$ is installed,
a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by some process will
set \var{B}.\fld{super} to a value read from \var{par}.\fld{head}.
We shall show that this happens soon enough that $B.\fld{super}$ can differ from the true index of $B'$
by at most 1.
%After $B$ is installed 
%in $\var{v}.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
%when they attempt to advance $\var{v}.\fld{head}$ from $h$ to $h+1$,
%using a value they read from the \fld{head} field of $\var{v}$'s parent.
%As mentioned above, this might not be the exact index of $B$'s superblock, but we
%shall prove that it is close.

@@

\opemph{Propagate}(\var{v}) guarantees that any blocks that are in \var{v}'s children when \op{Propagate} is invoked are propagated to the root.
It uses the double refresh idea described
above and invokes two \op{Refresh}es on \var{v} in Lines
\ref{firstRefresh} and \ref{secondRefresh}. 
If both fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
that propagated blocks that were in \var{v}'s children prior to line \ref{firstRefresh} to \var{v}.
Then, \op{Propagate} recurses to \var{v}.\fld{parent} to continue propagating blocks up to the root.

>>>>

\subsection{Overview}
Our \emph{\ordering\ tree} data structure is used to agree on a total ordering of the operations performed on the queue.
It is a static binary tree of height $\ceil{\log_2 p}$ with one leaf 
for each process. 
Each tree node  stores an array of \emph{blocks}, where each block represents a 
sequence of enqueues and a sequence of dequeues.
See Figure \ref{orderingtree} for an example.
In this section, we use an infinite array of blocks in each node.
Section \ref{reducing} describes how to replace the infinite array by a representation that uses bounded space.

@@

\Eric{Reworded following parag after Hossein's comments of Jan 4}
To propagate operations from a node $\var{v}$'s children to $\var{v}$, $P$ first observes
the blocks in both of $\var{v}$'s children that are not already in $\var{v}$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $\var{v}$'s \fld{blocks} array using a \op{CAS}.
Following \cite{DBLP:conf/fsttcs/JayantiP05}, we call this a 3-step sequence a
\op{Refresh} on $\var{v}$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on~$\var{v}$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $\var{v}$'s children to $\var{v}$,
we can prove that if two \op{Refresh}es by $P$ on $\var{v}$ fail,
then $P$'s operation has been propagated to $\var{v}$ by some other process, so $P$ can continue 
onwards towards the~root.

@@

\begin{figure*}[t]
\input{tournamentTree.pdf_t}
\caption{An example \ordering\ tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still propagating towards the root.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{orderingtree}}
\medskip
\input{implicit.pdf_t}
\caption{\label{implicit}The actual, implicit representation of the tree shown in Figure \ref{orderingtree}.
The leaf blocks simply show the \fld{element} field.
Internal blocks show the \fld{sum\sub{enq}} and \fld{sum\sub{deq}} fields,
and \eleft\ and \eright\ fields are shown using arrows as in Figure \ref{orderingtree}.
Root blocks also have the additional \fld{size} field.
The \fld{super} field is not shown.}
\end{figure*}

@@

Performing a double \op{Refresh} at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation completes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in the root's blocks array.
Thus, we linearize operations according to the block they belong to in the root's array.
We can choose how to order operations in the same block, since they must be concurrent.

@@

When linearizing the operations propagated to the root, we must
choose how to order operations within a block.  
We choose to put
each block's enqueues before its dequeues.
Thus, if the root's blocks array contains blocks $B_1, \ldots, B_k$, the 
linearization ordering~is 
\begin{equation}
L=E(B_1)\cdot D(B_1) \cdot E(B_2) \cdot D(B_2) \cdots E(B_k) \cdot D(B_k).
\label{linearization}
\end{equation}

@@

The only remaining field is $B.\fld{super}$.  When the block 
$B$ is created for a \node\ $\var{v}$, we do not yet know where its
superblock will eventually be installed in \var{v}'s parent.
So, we leave $B.\fld{super}$ blank.  
Soon after $B$ is installed,
%a call to \op{Advance} at line \ref{helpAdvance} or \ref{advance} by 
some process will
set \var{B}.\fld{super} to a value read from the \fld{head} field of \var{v}'s parent.
We shall show that this happens soon enough that $B.\fld{super}$ can differ from the true index of $B'$
by at most 1.
%After $B$ is installed 
%in $\var{v}.\fld{blocks}[\var{h}]$, processes cooperate to fill in $B.\fld{super}$ 
%when they attempt to advance $\var{v}.\fld{head}$ from $h$ to $h+1$,
%using a value they read from the \fld{head} field of $\var{v}$'s parent.
%As mentioned above, this might not be the exact index of $B$'s superblock, but we
%shall prove that it is close.

@@

\opemph{Propagate}(\var{v}) guarantees that any blocks that are in \var{v}'s children when \op{Propagate} is invoked are propagated to the root.
It uses the double \op{Refresh} idea described
above and invokes two \op{Refresh}es on \var{v} in Lines
\ref{firstRefresh} and \ref{secondRefresh}. 
If both fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
that propagated blocks that were in \var{v}'s children prior to line \ref{firstRefresh} to \var{v}.
Then, \op{Propagate} recurses to \var{v}.\fld{parent} to continue propagating blocks up to the root.

### 2023-05-11 podc_paper/analysis.tex
minor fixes

\begin{proposition}\label{casbound}
Each \op{Enqueue} or \op{Dequeue} operation performs $O(\log p)$ \op{CAS} instructions.
\end{proposition}
\begin{proof}
An operation invokes \op{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
A \op{Refresh} does at most 7 \op{CAS} steps: one in line \ref{cas} and two from each \op{Advance} in line \ref{helpAdvance} or~\ref{advance}.
\end{proof}

@@

The following lemma is useful in bounding the time for \op{GetEnqueue}
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by \Cref{lem::appendExactlyOnce}.
Thus, $B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

>>>>

\begin{proposition}\label{casbound}
Each \op{Enqueue} or \op{Dequeue} operation performs $O(\log p)$ \op{CAS} instructions.
\end{proposition}
\begin{proof}
An operation invokes \op{Refresh} at most twice at each of the $\ceil{\log_2 p}$ levels of the tree.
A \op{Refresh} does at most 5 \op{CAS} steps: one in line \ref{cas} and two during each \op{Advance} in line \ref{helpAdvance} or~\ref{advance}.
\end{proof}

@@

The following lemma helps bound the time for \op{GetEnqueue}.
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by \Cref{lem::appendExactlyOnce}.
Thus, $B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

### 2023-05-11 podc_paper/correctness-GC.tex
minor fixes

\begin{lemma}\label{RBTupdates}
Suppose a step of the algorithm changes $\var{v}.\fld{blocks}$ from a non-empty tree $T$ to $T'$.
If the set of \fld{index} values  in $T$ is $I$, then the set of \fld{index} values in $T'$ is 
$(I \cap [m-1,\infty))\cup \{\max(I) + 1\}$ for some $m$.
\end{lemma}
\begin{proof}
The RBT of a node is updated only at line \ref{appendLeafGC} or \ref{casGC}.

@@

If line \ref{casGC} of a \op{Refresh} modifies $\var{v}.\fld{blocks}$, then \var{v} is an internal node.
After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
then creates the  block \var{new},
and then calls \op{AddBlock}($T,new$) to create $T'$.
Line \ref{createIndexGC} sets $\var{new}.\var{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $new$ to it.
So the claim is satisfied.
\end{proof}

@@

\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
hold for the modified algorithm.  Their proofs are identical to those given in 
Section \ref{sec::basicProperties} since they  depend only on \Cref{lem::headProgress} (which can be replaced by \Cref{lem::headProgressGC})
and the definition of subblocks given in (\ref{defsubblock}).
In particular, \Cref{lem::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
values of \fld{sum\sub{enq}}  so the searches of a RBT based on
\fld{sum\sub{enq}} values in lines
\ref{getChildGC} and \ref{FRsearchGC} can be done in logarithmic time.

@@

Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{cas} 
installs $B'$ in $\var{v}.\fld{blocks}$.
The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in \var{v} contains $B$ as a subblock
and therefore \var{v} contains $op$.

>>>>

\begin{lemma}\label{RBTupdates}
Suppose a step of the algorithm changes $\var{v}.\fld{blocks}$ from a non-empty tree $T$ to $T'$.
If the set of \fld{index} values  in $T$ is $I$, then the set of \fld{index} values in $T'$ is 
$(I \cap [m-1,\infty))\cup \{\max(I) + 1\}$ for some~$m$.
\end{lemma}
\begin{proof}
The RBT of a node is updated only at line \ref{appendLeafGC} or \ref{casGC}.

@@

If line \ref{casGC} of a \op{Refresh} modifies $\var{v}.\fld{blocks}$, then \var{v} is an internal node.
After reading $T$ from $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC},
the \op{Refresh} then creates the  block \var{new},
and then calls \op{AddBlock}($T,new$) to create $T'$.
Line \ref{createIndexGC} sets $\var{new}.\var{index} = \max(I)+1$.
The \op{AddBlock} that creates $T'$ may optionally \op{Split} the RBT and then add $new$ to it.
So the claim is satisfied.
\end{proof}

@@

\Cref{lem::subblocksDistinct}, \Cref{lem::noDuplicates}, \Cref{lem::sum}, \Cref{blockNotEmpty} 
hold for the modified algorithm.  Their proofs are identical to those given in 
Section \ref{sec::basicProperties} since they  depend only on \Cref{lem::headProgress} (which can be replaced by \Cref{lem::headProgressGC})
and the definition of subblocks given in (\ref{defsubblock}).
In particular, \Cref{lem::sum} says that nodes in an in-order traversal of a RBT have non-decreasing
values of \fld{sum\sub{enq}}  so the searches of a RBT based on
\fld{sum\sub{enq}} values in lines
\ref{FRsearchGC} and \ref{getChildGC} can be done in logarithmic time.

@@

Suppose $R$'s call to \op{CreateBlock} returns a new block $B'$.  Then, $R$'s \op{CAS} at line \ref{casGC} 
installs $B'$ in $\var{v}.\fld{blocks}$.
The \op{CreateBlock} set $B'.\eleft$ to the maximum \fld{index} in $\var{v}.\fld{left}$'s RBT at line \ref{createEndLeftGC}.
By \Cref{nonEmptyIncreasing}, this maximum \fld{index} is bigger than $i$.
By the definition of subblocks, some block in \var{v} contains $B$ as a subblock
and therefore \var{v} contains $op$.

### 2023-05-11 podc_paper/introduction.tex
minor fixes

There has been a great deal of research in the past several decades on the design of shared queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management \cite{BBFRSW21}\here{try to find an older, more canonical reference},
packet processing \cite{DPDK}, synchronization \cite{KAE23},\here{is this a good citation for this?}
and sharing resources or tasks.
We focus on shared queues that are \emph{linearizable} \cite{HW90}, meaning that operations
appear to take place atomically, and \emph{lock-free}, meaning that some operation on the queue
is guaranteed to complete regardless of how asynchronous processes are scheduled to take~steps.

@@

We exploit this loophole.  We show  it is, in fact, possible for a linearizable queue
to have step complexity sublinear in $p$.
Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
It is \emph{wait-free}, meaning that every operation is guaranteed to complete within a finite number of its own steps.
For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
dequeues take $O(\log^2 p + \log q)$ steps,
and then modify it to bound the space
while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
Both versions of our queue use single-word CAS on reasonably-sized 
words.
We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored in $O(1)$ words.
This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.
For the space-bounded version, we unlink unneeded objects from our data structure.
We do not address the orthogonal problem of reclaiming memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

>>>>

There has been a great deal of research in the past several decades on the design of shared queues.
Besides being a fundamental data structure, queues are used in
significant concurrent applications, including OS kernels \cite{MP91}, memory management \cite{BBFRSW21}\here{try to find an older, more canonical reference},
%packet processing \cite{DPDK}, 
synchronization \cite{KAE23},\here{is this a good citation for this?}
and sharing resources or tasks.
We focus on shared queues that are \emph{linearizable} \cite{HW90}, meaning that operations
appear to take place atomically, and \emph{lock-free}, meaning that some operation on the queue
is guaranteed to complete regardless of how asynchronous processes are scheduled to take~steps.

@@

We exploit this loophole.  We show  it is, in fact, possible for a linearizable queue
to have step complexity sublinear in $p$.
Our queue is the first whose step complexity  is polylogarithmic in $p$ and in $q$, the number of elements in the queue.
It is \emph{wait-free}, meaning that every operation is guaranteed to complete within a finite number of its own steps.
For ease of presentation, we first give an unbounded-space construction where enqueues take $O(\log p)$ steps and
dequeues take $O(\log^2 p + \log q)$ steps,
and then modify it to bound the space
while  having $O(\log p\log( p+ q))$ amortized step complexity  per operation.
Moreover, each operation does $O(\log p)$ CAS instructions in the worst case, whereas previous
lock-free queues use 
%an unbounded number of CAS instructions in the worst case and 
$\Omega(p)$ CAS instructions, even in an amortized sense.
Both versions of our queue use single-word CAS on reasonably-sized 
words.
We assume that a word is large enough to store an item to be enqueued (or at least a pointer to it).  We also assume that the number of operations performed on the queue can be stored (in binary) in $O(1)$ words.
This is analogous to the assumption for the classical RAM model that the number of bits per word is logarithmic in the problem size.
For the space-bounded version, we unlink unneeded objects from our data structure.
We do not address the orthogonal problem of reclaiming memory; we assume a safe
garbage collector, such as the highly optimized one that Java provides.

### 2023-05-11 podc_paper/related.tex
minor fixes

Many papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty.
% or when concurrent operations can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduces contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
Enqueues within a basket order  themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.

@@

\paragraph{Array-Based Queues.}
Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
%Similar to the front and back pointers of list-based queues, indices
Dequeues and enqueues update
indexes of the front and back elements using CAS instructions.
Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
the capacity constraint by using a linked list of arrays.
These solutions also use $\Omega(p)$ steps per operation due to the CAS retry problem.

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
(It was shown recently that this implementation can be modified to use single-word CAS instructions rather than double-width CAS \cite{RK23}.)
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
Other queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} also used this methodology
on array-based queues.
In worst-case executions that use the slow path,
they also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.

>>>>

Many papers have described ways to reduce contention in the MS-queue.
Moir et al.~\cite{DBLP:conf/spaa/MoirNSS05} 
added an elimination array that allows an enqueue to pass its enqueued value directly
to a concurrent dequeue when the queue is empty.
% or when concurrent operations can be linearized to empty the queue.
However, when there are $p$ concurrent enqueues (and no dequeues), the CAS retry problem
is still present.
The baskets queue of
Hoffman, Shalev, and Shavit~\cite{DBLP:conf/opodis/HoffmanSS07} 
attempts to reduce contention by grouping concurrent enqueues into baskets.
An enqueue that fails its CAS is put in the basket with the enqueue that succeeded.
Enqueues within a basket order  themselves without having to access the back of the queue.
However, if $p$ concurrent enqueues are in the same basket
the CAS retry problem occurs when they order themselves using CAS instructions.
Both modifications still have $\Omega(p)$ amortized step complexity.

@@

\paragraph{Array-Based Queues.}
Arrays can be used to implement queues with bounded capacity \cite{DBLP:conf/iceccs/ColvinG05,DBLP:conf/icdcn/Shafiei09,DBLP:conf/spaa/TsigasZ01}.  
%Similar to the front and back pointers of list-based queues, indices
Dequeues and enqueues update
indices of the front and back elements using CAS instructions.
Gidenstam, Sundell, and Tsigas~\cite{DBLP:conf/opodis/GidenstamST10} avoid
the capacity constraint by using a linked list of arrays.
These solutions also use $\Omega(p)$ steps per operation due to the CAS retry problem.

Morrison and Afek \cite{DBLP:conf/ppopp/MorrisonA13} also used a linked list of (circular) arrays.
To avoid the CAS retry problem, concurrent operations try to claim spots in an array using fetch\&increment instructions.
(It was shown recently that this implementation can be modified to use single-word CAS instructions rather than double-width CAS \cite{RK23}.)
If livelock between enqueues and a racing dequeue prevent enqueues from claiming a spot,
the enqueues fall back on using a CAS to add a new array to the linked list, 
and the CAS retry problem reappears.
% I looked at the CRQ algorithm to implement each of the circular arrays.
% Here is an execution that causes livelock for 3 processes
% enqueue(A) reads t=0 and increments tail to 1
% enqueue(B) reads t=1 and increments tail to 2
% (*) dequeue reads h=0 and increments head to 1
% dequeue's CAS succeds changing array[0] to (1,R,bottom)
% dequeue reads t=2 at line 53
% enqueue(A) fails CAS on array[0]
% dequeue reads h=1 and increments head to 2
% dequeue's CAS succeeds changing array[1] to (1,R+1,bottom)
% enqueue(B) fails CAS on array[1]
% enqueue(A) reads t=2 and increments tail to 3
% enqueue(B) reads t=3 and increments tail to 4
% dequeue reads t=4 at line 53
% go back to (*) and repeat this sequence
This approach is similar to the fast-path slow-path methodology \cite{10.1145/2370036.2145835}.
Other array-based queues \cite{Nik19,10.1145/3490148.3538572,DBLP:conf/ppopp/YangM16} also used this methodology.
In worst-case executions that use the slow path,
they also take $\Omega(p)$ steps per operation, 
due either to the CAS retry problem or helping mechanisms.
